---
title: "Flocks, Herds, Swarms, and Schools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Flocks, Herds, Swarms, and Schools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

The same boids rules can be tuned to read as different collective-motion
patterns. This vignette uses 3D examples as the main view, then adds 2D
overhead variants where they help explain the movement.

```{r}
library(boids4R)
```

## Helpers

```{r}
final_frame <- function(sim) {
  frames <- as.data.frame(sim)
  frames[frames$frame == max(frames$frame), , drop = FALSE]
}

mean_nearest_neighbor <- function(frame) {
  if (nrow(frame) < 2L) return(NA_real_)
  coords <- as.matrix(frame[, c("x", "y", "z"), drop = FALSE])
  distances <- as.matrix(stats::dist(coords))
  diag(distances) <- NA_real_
  mean(apply(distances, 1L, min, na.rm = TRUE), na.rm = TRUE)
}

movement_summary <- function(sim, label) {
  frames <- as.data.frame(sim)
  final <- final_frame(sim)
  data.frame(
    label = label,
    dimension = sim$dimension,
    boids = length(unique(final$id)),
    species = paste(sort(unique(final$species)), collapse = ", "),
    frames = length(unique(frames$frame)),
    mean_speed = round(mean(final$speed), 3),
    xy_spread = round(mean(sqrt((final$x - mean(final$x))^2 + (final$y - mean(final$y))^2)), 3),
    z_spread = round(stats::sd(final$z), 3),
    mean_nearest_neighbor = round(mean_nearest_neighbor(final), 3),
    stringsAsFactors = FALSE
  )
}

species_palette <- function(species) {
  keys <- sort(unique(species))
  stats::setNames(grDevices::hcl.colors(length(keys), "Dark 3"), keys)
}

draw_projection <- function(sim, title, x_axis = "x", y_axis = "y") {
  final <- final_frame(sim)
  world <- sim$world
  palette <- species_palette(final$species)
  xlim <- if (x_axis %in% rownames(world$bounds)) world$bounds[x_axis, ] else range(final[[x_axis]])
  ylim <- if (y_axis %in% rownames(world$bounds)) world$bounds[y_axis, ] else range(final[[y_axis]])

  graphics::plot(
    final[[x_axis]], final[[y_axis]],
    xlim = xlim,
    ylim = ylim,
    asp = 1,
    xlab = x_axis,
    ylab = y_axis,
    main = title,
    col = palette[final$species],
    pch = 16,
    cex = 0.7
  )
  graphics::legend("topright", legend = names(palette), col = palette, pch = 16, bty = "n", cex = 0.75)
}

draw_two_projections <- function(sim, title) {
  old_par <- graphics::par(mfrow = c(1, 2), mar = c(3, 3, 3, 1))
  draw_projection(sim, paste(title, "x/y"), "x", "y")
  draw_projection(sim, paste(title, "x/z"), "x", "z")
  graphics::par(old_par)
}
```

## Build example simulations

Flocks and swarms use the named 3D scenarios. The school example narrows the
3D bounds into a water-column shape. The herd example is also 3D, but with a
shallow vertical extent to represent animals moving over uneven ground.

```{r}
flock_3d <- boids_scenario(
  "murmuration_3d",
  n = 180,
  steps = 70,
  record_every = 5,
  seed = 501
)

swarm_3d <- boids_scenario(
  "mixed_species_3d",
  n = 180,
  steps = 70,
  record_every = 5,
  seed = 502
)

school_bounds <- matrix(
  c(-2.2, -1.25, -0.7, 2.2, 1.25, 0.7),
  ncol = 2,
  dimnames = list(c("x", "y", "z"), c("min", "max"))
)
school_3d <- simulate_boids(
  boids_state(170, "3d", bounds = school_bounds, seed = 503),
  boids_world(
    "3d",
    bounds = school_bounds,
    boundary = "wrap",
    attractors = data.frame(x = 0.75, y = -0.15, z = 0.05, strength = 0.32)
  ),
  boids_params(
    "3d",
    separation_weight = 1.20,
    alignment_weight = 1.15,
    cohesion_weight = 0.98,
    cohesion_radius = 0.72,
    alignment_radius = 0.55,
    max_speed = 1.20,
    noise = 0.001
  ),
  steps = 70,
  record_every = 5,
  seed = 504
)

herd_bounds <- matrix(
  c(-2.4, -1.35, -0.08, 2.4, 1.35, 0.08),
  ncol = 2,
  dimnames = list(c("x", "y", "z"), c("min", "max"))
)
set.seed(505)
herd_positions <- cbind(
  stats::runif(150, -2.15, -1.25),
  stats::runif(150, -0.55, 0.65),
  stats::rnorm(150, 0, 0.015)
)
herd_velocities <- cbind(
  stats::runif(150, 0.10, 0.42),
  stats::rnorm(150, 0, 0.08),
  stats::rnorm(150, 0, 0.005)
)
herd_3d <- simulate_boids(
  boids_state(
    150,
    "3d",
    bounds = herd_bounds,
    positions = herd_positions,
    velocities = herd_velocities,
    species = rep(c("lead", "middle", "edge"), length.out = 150),
    seed = 505
  ),
  boids_world(
    "3d",
    bounds = herd_bounds,
    boundary = "reflect",
    predators = data.frame(x = -1.75, y = 0.95, z = 0, radius = 0.72, strength = 0.9),
    attractors = data.frame(x = 2.0, y = -0.45, z = 0, strength = 0.55)
  ),
  boids_params(
    "3d",
    separation_weight = 1.05,
    alignment_weight = 0.92,
    cohesion_weight = 0.86,
    predator_weight = 2.4,
    goal_weight = 0.18,
    max_speed = 1.05,
    max_force = 0.095,
    noise = 0.0005
  ),
  steps = 75,
  record_every = 5,
  seed = 506
)
```

## Compare the 3D examples

```{r}
examples_3d <- list(
  flock = flock_3d,
  herd = herd_3d,
  swarm = swarm_3d,
  school = school_3d
)

do.call(rbind, Map(movement_summary, examples_3d, names(examples_3d)))
```

The x/y view shows the collective shape from above. The x/z view reveals which
examples use a full 3D volume and which stay near a ground or water layer.

```{r flock-3d-projection, fig.width = 8, fig.height = 4}
draw_two_projections(flock_3d, "flock")
```

```{r herd-3d-projection, fig.width = 8, fig.height = 4}
draw_two_projections(herd_3d, "herd")
```

```{r swarm-3d-projection, fig.width = 8, fig.height = 4}
draw_two_projections(swarm_3d, "swarm")
```

```{r school-3d-projection, fig.width = 8, fig.height = 4}
draw_two_projections(school_3d, "school")
```

## 2D variants

Overhead 2D examples are useful for corridor, schooling, and avoidance
experiments where the top-down geometry is the main story.

```{r}
flock_2d <- boids_scenario(
  "schooling_2d",
  n = 130,
  steps = 60,
  record_every = 5,
  seed = 601
)

herd_2d <- boids_scenario(
  "predator_avoidance_2d",
  n = 130,
  steps = 65,
  record_every = 5,
  seed = 602
)

school_2d <- boids_scenario(
  "obstacle_corridor_2d",
  n = 130,
  steps = 65,
  record_every = 5,
  seed = 603
)

examples_2d <- list(
  top_down_flock = flock_2d,
  avoidance_herd = herd_2d,
  obstacle_school = school_2d
)

do.call(rbind, Map(movement_summary, examples_2d, names(examples_2d)))
```

```{r two-d-variants, fig.width = 8, fig.height = 8}
old_par <- graphics::par(mfrow = c(2, 2), mar = c(3, 3, 3, 1))
draw_projection(flock_2d, "2D top-down flock", "x", "y")
draw_projection(herd_2d, "2D avoidance herd", "x", "y")
draw_projection(school_2d, "2D obstacle school", "x", "y")
graphics::par(old_par)
```

## Animate with ggWebGL

When `ggWebGL` 0.4.0 or later is installed, any of these simulations can be
handed to the optional adapter for timeline animation.

```{r eval = FALSE}
if (requireNamespace("ggWebGL", quietly = TRUE) &&
    utils::packageVersion("ggWebGL") >= "0.4.0") {
  spec <- as_ggwebgl_spec(flock_3d, vector_every = 14, shader = "density_splat")
  spec$render$timeline$autoplay <- TRUE
  ggWebGL::ggWebGL(spec, height = 540)
}
```
