---
title: "Getting Started with boids4R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with boids4R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`boids4R` simulates flocking and swarm dynamics in R. The core objects are
renderer-neutral: they describe boids, rules, worlds, and recorded frames.

```{r}
library(boids4R)

sim <- boids_scenario("schooling_2d", n = 80, steps = 20, seed = 12)
frames <- as.data.frame(sim)
head(frames)
```

The same frame table can be handed to visualization packages. If `ggWebGL` is
installed, the optional adapter creates point and velocity-vector primitives
with exact timeline controls.

```{r eval = FALSE}
if (requireNamespace("ggWebGL", quietly = TRUE)) {
  ggWebGL::ggWebGL(as_ggwebgl_spec(sim), height = 440)
}
```

For larger examples, see the scenario gallery and custom simulation workflow
vignettes. They show obstacle corridors, predator avoidance, parameter sweeps,
and mixed-species 3D runs using the same renderer-neutral frame output.

To generate a standalone WebGL page for an external browser:
```{r eval = FALSE}
sim <- boids4R::boids_scenario("murmuration_3d", n = 400, steps = 150, seed = 1)

spec <- boids4R::as_ggwebgl_spec(sim)
spec$render$timeline$autoplay <- TRUE
spec$render$timeline$speed <- 2

w <- ggWebGL::ggWebGL(spec, height = 520)

htmlwidgets::saveWidget(w, "boids_murmuration.html", selfcontained = FALSE)
browseURL(normalizePath("boids_murmuration.html"))
```

To see trajectories, not only moving current positions, use cumulative line trails:
```{r eval = FALSE}
frames <- as.data.frame(sim)
keep <- unique(frames$id)[1:120]
trail <- frames[frames$id %in% keep, ]

line_layer <- ggWebGL::ggwebgl_layer_lines(
  trail,
  x = "x", y = "y", z = "z",
  group = "id",
  colour = "#334155",
  alpha = 0.08,
  width = 0.7,
  frame = "frame",
  time = "time"
)

point_layer <- ggWebGL::ggwebgl_layer_points(
  frames,
  x = "x", y = "y", z = "z",
  colour = "#2563eb",
  alpha = 0.45,
  size = 2,
  id = "id",
  frame = "frame",
  time = "time"
)

spec <- ggWebGL::ggwebgl_spec(
  list(line_layer, point_layer),
  webgl = list(
    view = ggWebGL::ggwebgl_view("3d", controller = "orbit", projection = "perspective")
  ),
  timeline = ggWebGL::ggwebgl_timeline(
    frames = sort(unique(frames$frame)),
    filter = "cumulative",
    autoplay = TRUE,
    speed = 2
  )
)

htmlwidgets::saveWidget(ggWebGL::ggWebGL(spec, height = 520), "boids_trails.html")
browseURL(normalizePath("boids_trails.html"))
```
