---
title: "Swarm Art Creation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Swarm Art Creation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

Recorded boids frames can be used as generative drawing material. This vignette
keeps the simulation renderer-neutral, then turns frame tables into static
artworks with base R graphics.

```{r}
library(boids4R)
```

## Helpers

```{r}
frame_table <- function(sim) {
  frames <- as.data.frame(sim)
  frames[order(frames$id, frames$frame), , drop = FALSE]
}

world_limits <- function(sim) {
  list(
    xlim = sim$world$bounds["x", ],
    ylim = sim$world$bounds["y", ]
  )
}

draw_empty_canvas <- function(sim, title = "") {
  lim <- world_limits(sim)
  graphics::plot(
    NA_real_, NA_real_,
    xlim = lim$xlim,
    ylim = lim$ylim,
    asp = 1,
    axes = FALSE,
    xlab = "",
    ylab = "",
    main = title
  )
}

fade_palette <- function(n, palette = "Inferno") {
  grDevices::hcl.colors(n, palette)
}
```

## Trail drawing

Line trails turn motion into a dense drawing. The example below draws only a
subset of boids so individual paths remain visible.

```{r}
trail_sim <- boids_scenario(
  "murmuration_3d",
  n = 140,
  steps = 95,
  record_every = 2,
  seed = 710
)

trail_frames <- frame_table(trail_sim)
keep_ids <- unique(trail_frames$id)[seq(1, length(unique(trail_frames$id)), by = 3)]
trail_frames <- trail_frames[trail_frames$id %in% keep_ids, , drop = FALSE]
```

```{r trail-art, fig.width = 7, fig.height = 7}
draw_empty_canvas(trail_sim, "murmuration trails")
ids <- unique(trail_frames$id)
cols <- grDevices::adjustcolor(fade_palette(length(ids), "Dark 3"), alpha.f = 0.22)
for (i in seq_along(ids)) {
  path <- trail_frames[trail_frames$id == ids[i], , drop = FALSE]
  graphics::lines(path$x, path$y, col = cols[i], lwd = 0.8)
}
```

## Time-layered particles

A different style keeps all boids but draws successive frames with increasing
opacity and size. Recent frames become the bright foreground.

```{r}
particle_sim <- boids_scenario(
  "schooling_2d",
  n = 180,
  steps = 75,
  record_every = 3,
  seed = 720
)
particle_frames <- as.data.frame(particle_sim)
frames <- sort(unique(particle_frames$frame))
```

```{r particle-art, fig.width = 7, fig.height = 7}
draw_empty_canvas(particle_sim, "time-layered school")
frame_cols <- vapply(
  seq_along(frames),
  function(i) {
    grDevices::adjustcolor(
      fade_palette(length(frames), "Viridis")[i],
      alpha.f = seq(0.06, 0.55, length.out = length(frames))[i]
    )
  },
  character(1)
)
for (i in seq_along(frames)) {
  layer <- particle_frames[particle_frames$frame == frames[i], , drop = FALSE]
  graphics::points(layer$x, layer$y, pch = 16, cex = 0.25 + 0.45 * i / length(frames), col = frame_cols[i])
}
```

## Negative-space obstacles

Obstacle and predator avoidance can produce visual gaps. Here the obstacles are
drawn as quiet negative-space forms under the flock traces.

```{r}
negative_sim <- boids_scenario(
  "obstacle_corridor_2d",
  n = 170,
  steps = 85,
  record_every = 3,
  seed = 730
)
negative_frames <- frame_table(negative_sim)
negative_ids <- unique(negative_frames$id)[seq(1, length(unique(negative_frames$id)), by = 2)]
negative_frames <- negative_frames[negative_frames$id %in% negative_ids, , drop = FALSE]
```

```{r negative-space-art, fig.width = 7, fig.height = 5}
draw_empty_canvas(negative_sim, "negative-space corridor")
for (i in seq_len(nrow(negative_sim$world$obstacles))) {
  graphics::symbols(
    negative_sim$world$obstacles$x[i],
    negative_sim$world$obstacles$y[i],
    circles = negative_sim$world$obstacles$radius[i],
    inches = FALSE,
    add = TRUE,
    bg = "white",
    fg = "gray85"
  )
}
cols <- grDevices::adjustcolor(fade_palette(length(negative_ids), "Plasma"), alpha.f = 0.18)
for (i in seq_along(negative_ids)) {
  path <- negative_frames[negative_frames$id == negative_ids[i], , drop = FALSE]
  graphics::lines(path$x, path$y, col = cols[i], lwd = 0.9)
}
```

## Depth as colour

For 3D simulations, z can drive colour or point size in a 2D projection. This
creates a depth print without needing a 3D renderer.

```{r}
depth_sim <- boids_scenario(
  "mixed_species_3d",
  n = 190,
  steps = 70,
  record_every = 5,
  seed = 740
)
depth_final <- as.data.frame(depth_sim)
depth_final <- depth_final[depth_final$frame == max(depth_final$frame), , drop = FALSE]
depth_rank <- (depth_final$z - min(depth_final$z)) / diff(range(depth_final$z))
```

```{r depth-art, fig.width = 7, fig.height = 7}
draw_empty_canvas(depth_sim, "3D depth print")
depth_cols <- fade_palette(100, "BluYl")
graphics::points(
  depth_final$x,
  depth_final$y,
  pch = 16,
  cex = 0.35 + 0.9 * depth_rank,
  col = grDevices::adjustcolor(depth_cols[pmax(1, ceiling(depth_rank * 99))], alpha.f = 0.7)
)
```

## Exporting art

Use normal graphics devices to export a static artwork. The same frame table can
also be sent to `ggWebGL` when an animated artwork is preferable.

```{r eval = FALSE}
png("swarm-art.png", width = 1800, height = 1800, res = 220)
draw_empty_canvas(trail_sim, "murmuration trails")
for (i in seq_along(ids)) {
  path <- trail_frames[trail_frames$id == ids[i], , drop = FALSE]
  graphics::lines(path$x, path$y, col = cols[i], lwd = 0.8)
}
dev.off()
```

```{r eval = FALSE}
if (requireNamespace("ggWebGL", quietly = TRUE) &&
    utils::packageVersion("ggWebGL") >= "0.4.0") {
  spec <- as_ggwebgl_spec(depth_sim, vector_every = 18, shader = "density_splat")
  spec$render$timeline$autoplay <- TRUE
  ggWebGL::ggWebGL(spec, height = 540)
}
```
